---
title: "Case Study Version 2"
author: "McKenna Weech"
date: "10/8/2019"
output: 
  github_document:
    pandoc_args: --webtex
---

This case study is largely baised off of a case study made for StanCon2019 which can be found here https://github.com/lauken13/Beginners_Bayes_Workshop

Stan Beginners Guide
================

This guide will introduce first time Stan users to Stan inlcuding what it
does, how it works, what it looks like, and how to begin using it.

-----

#### What is Stan and why do we need it?

Stan is a programming languge that helps us to run larger and more
complex beysian models. Without Stan it would be impossible to run many
of the models that we would like to use.

#### What does Stan do?

Stan works by running complex Marcov chain Monte Carlo algorithums to calculate the log probability density function.  

#### What does Stan code look like?

There are three required blocks in Stan code:

`Data`

`Parameter`

`Model`

The **Data** block is used to declare the data that we will use in the model. 

The **Parameter** block is used to specify the parameters. 

The **Model** block is used to define the model including the likelihood and priors. 

#### How do you start using Stan?

Stan runs in the backgroud of r, so you will be working with both an r markdown or r scrip file as well as a stan file. Start with the r file where you have your data. You will then have to open an r stan file which is another file type option in r studio. 

When you first open a Stan file it should look like this: 
```{
//
// This Stan program defines a simple model, with a
// vector of values 'y' modeled as normally distributed
// with mean 'mu' and standard deviation 'sigma'.
//
// Learn more about model development with Stan at:
//
//    http://mc-stan.org/users/interfaces/rstan.html
//    https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started
//

// The input data is a vector 'y' of length 'N'.
data {
  int<lower=0> N;
  vector[N] y;
}

// The parameters accepted by the model. Our model
// accepts two parameters 'mu' and 'sigma'.
parameters {
  real mu;
  real<lower=0> sigma;
}

// The model to be estimated. We model the output
// 'y' to be normally distributed with mean 'mu'
// and standard deviation 'sigma'.
model {
  y ~ normal(mu, sigma);
}
```
Here you can see the three required stan blocks along with some preset basic model options and some comments and information about stan. 

Now we will look at each block one at a time.

### Data Block 
```{
data {
  int<lower=0> N;
  vector[N] y;
}
```
First, you can see that a block is denoted with { } brackets and that after every line of code there is a ; semi-colon. These are necessary for the code to run. 

As we said before, the data block is where you declare your data. Your actual data will be in your r file but this tells stan what to look for. For every element, you have to declare what type of data it is.

There are two basic types of data you can declare ```int``` or ```real```. 

**int** = count data 

**real** = continuious data 

There are other more advanced options for declaring dating using vectors and matrices which can be a more effecient ways to declare data and may be worth looking into for more complex model and data types. 

It is also worth noting that stan is vectorized for effeciency. 

Another thing you will notice is ```N```. Stan requires you to declare how my observaions you have as a variable N. 

Next you need to declare any variables in your model and their type. In the example above there is one dependent variable ```y``` declared in the data. ```y``` is declared as an array with ```N``` observations. Again there are multiple ways to declare data. For example, this could have been declared with ```vector[N] y``` for essentially the same result.  

```<lower=1>``` is just a saftey measure that we can use to make sure that the data we insert is in the right form. For example, if you were to specify ```<lower=1>``` Stan would stop and give a warning message if you try to pass it any data with N less than 1, but this is not nessecary and is just a saftey check if you need it. 

### Parameter Block 
```
parameters {
  real mu;
  real<lower=0> sigma;
}
```

The parameters block is similar to the data block but here you will declare your parameters. Just like in the data block you will have to declare what type of data they are. You can see that here they suggest using the saftey check ```<lower=0>``` on sigma becasue you can't have a negative sigma. 

### Model Block 
```
model {
  y ~ normal(mu, sigma);
}
```
The model block is where you will specify your model and your priors. This is just an example of a basic regression model. 


---

Stan Case Study 
===============

```{r}
# Load libraries. 

library(tidyverse)
library(rstan)
library(bayesplot)
```

```{r load-data}
# Import data. 

store_data <- read_csv('data/store_data.csv')
store <- as.tibble(store_data)
store

```
### Background 

Imagine that you are a consultant working at a firm in Salt Lake City. A large retail chain baised in SLC approches you to hire you for a job. The manager explains to you that they are worried about the number of complaints they receive from upset customers who feel like they did not receive enough employees assistance at one of their stores. Their curent policy is to staff two employees at every store at all times. While this is the current company policy, the manager believes that two attendents could be too many for some of their locations and not enough for others and is an inefficient allocatiton of company resouces. 

One alternative to this policy is to vary the number of employees from store to store depending on the demand of the customers. 

### The Goal 

The manager wishes to employ your services to help them to find the optimal
number of employees they should place in each of their stores in
order to minimize the number of complaints while also keeping
expenditure on employess affordable.

## Building a Model 


Because you are all pros in bayesian statistics you begin by thinking about you want to model this type of data. A poisson distribution doesn't allow for any negative amount of complaints which would make sense because people can't call a negative amount of times. 

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t})} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{employees}_{b,t}
\end{align*}
$$

### Prior Predictive Check 
Now we want to run a prior predictive check to see what the priors we have choosen are saying is possible. 

```{r, simulate-poisson-data, eval = FALSE }
# using normal distributions for priors on alpha and beta 
prior_predictive <- function(employees, alpha_mean, alpha_sd, beta_mean, beta_sd) {
  N <- length(employees)
  alpha <- rnorm(1, mean = alpha_mean, sd = alpha_sd);
  beta <- rnorm(1, mean = beta_mean, sd = beta_sd);
  complaints <- rpois(N, lambda = exp(alpha + beta * employees))
  return(complaints)
}
```

```{r, prior-pred-really-bad, eval = FALSE }
# you can run this chunk multiple times to keep generating different datasets
prior_predictive(
  store$employees,
  alpha_mean = 0,
  alpha_sd = 10,
  beta_mean = 0,
  beta_sd = 10
)
```

### Exploratory Data Analysis (Modeling) (STAN!)

Now we get to really run the stan code. Try following the steps listed in the beginning to write stan code. You can also look at my Stan file titled version_2_stan in the stan_programs folder.

Once we have writen the Stan we need to compile the model. 

```{r, eval = FALSE }
comp_model <- stan_model('stan_programs/version_2_stan.stan')
```

The Stan model is now compiled so we need to prepare the data to be passed to the model. We have already uploaded our data into R but Stan requires that it is in a specific format. 

We input data into stan in a list format. The things in your list need to match the data you declared in Stan. 
```{r, eval = FALSE }
standata_simple <- list(
  N = nrow(store), 
  complaints = store$complaints,
  employees = store$employees
)
str(standata_simple)
```

### Smapling

Here we are going to bring our complied stan model and our data together and start to look at what we've got. 
```{r, eval = FALSE }
fit_simple <- sampling(comp_model, data = standata_simple,
                       # these are the defaults but specifying them anyway
                       # so you can see how to use them: 
                       # posterior sample size = chains * (iter-warmup)
                       chains = 4, iter = 2000, warmup = 1000)
```


```{r, eval = FALSE }
print(fit_simple, pars = c('alpha','beta'))
```
```{r, mcmc_hist, eval = FALSE }
# https://mc-stan.org/bayesplot/reference/MCMC-distributions
#draws <- as.matrix(fit_simple, pars = c('alpha','beta'))
#draws_tibble <- as.tibble(draws)
# ggplot(draws_tibble, aes(alpha)) +    # Marginal posteriors of alpha.
#   geom_histogram()
#
# ggplot(draws_tibble, aes(beta)) +     # Marginal posteriors of beta.
#   geom_histogram()
#
# ggplot(draws_tibble, aes(x = alpha, y = beta)) +  # Posterior of (alpha,beta).
#   geom_jitter()

draws <- as.matrix(fit_simple, pars = c('alpha','beta'))
head(draws)
mcmc_hist(draws) # marginal posteriors of alpha and beta
mcmc_scatter(draws, alpha = 0.2, size = 1) # posterior of (alpha,beta)

```
```{r, compare-prior-posterior, eval = FALSE }
alpha_prior_post <- cbind(alpha_prior = rnorm(4000, log(7), 1), 
                          alpha_posterior = draws[, "alpha"])
mcmc_hist(alpha_prior_post, facet_args = list(nrow = 2), binwidth = 0.1) + 
  xlim(range(alpha_prior_post))


beta_prior_post <- cbind(beta_prior = rnorm(4000, -0.25, 0.5), 
                         beta_posterior = draws[, "beta"])
mcmc_hist(beta_prior_post, facet_args = list(nrow = 2), binwidth = 0.05) + 
  xlim(range(beta_prior_post))
```
### Posterior Predictive Check 

```{r, eval = FALSE }
fit_summary <- summary(fit_simple)
print(names(fit_summary))

print(fit_summary$summary)
```
```{r, eval = FALSE }
alpha_beta_summary <- summary(fit_simple, pars = c("alpha", "beta"), probs = c(0.1, 0.9))$summary
print(alpha_beta_summary)
```


### Model comparison
If we were really creating a model we would want to iterate through this a few times and then compare our models using one of the method that we have talked about in class. 

### Decision Analysis 
